---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# faustutils

<!-- badges: start -->
<!-- badges: end -->

The goal of faustutils is to provide convenient functions to interrogate a FAUST clustering and extract output from it.

## Installation

You can install the package from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("MiguelRodo/faustutils")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(faustutils)
```

First save the directory FAUST saved results to:
```{r }
# set the FAUST project path
proj_path <- usethis::proj_path('tests/testthat')
```

You can get the names of the markers FAUST used to cluster and the number of levels for each (after FAUST has clustered):
```{r }
get_faust_markers_and_levels(project_path = proj_path)
```

You can save FAUST-identified clusters as FCS files using only a subset of the markers (not run):
```{r , eval = FALSE}
save_faust_pop(project_path = proj_path, pop = list("CD3" = 2, "CD4" = 2), 
               gs = gs) # set gs to be the GatingSet FAUST clustered
```

You can get a table of counts of FAUST-identified clusters defined
using only a subset of the markers used (not run):
```{r , eval = FALSE}
get_pop_counts(project_path = proj_path, 
               pop = c("CD4" = "-", "CD8" = "+"))
```

You can save plots of counts of FAUST-identified clusters, where the clusters are defined using only a subset of the markers (not run):
```{r, eval = FALSE}
# plot all subsets that match these annotations individually
pop <- c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-")
plot_faust_count(project_path = proj_path,
                 pop = pop)

# plot total counts of subsets that match these annotations
# within each list element
pop <- list(c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-"),
           c("CD3" = "+", "CD4" = "-", "CD8-IgD" = "+"))
plot_faust_count(project_path = proj_path,
                 pop = pop)
```
