---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# UtilsFaustSV

<!-- badges: start -->
[![R-CMD-check](https://github.com/SATVILab/UtilsFaustSV/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/SATVILab/UtilsFaustSV/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/SATVILab/UtilsFaustSV/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/SATVILab/UtilsFaustSV/actions/workflows/test-coverage.yaml)
[![codecov](https://codecov.io/gh/SATVILab/UtilsFaustSV/branch/main/graph/badge.svg)](https://codecov.io/gh/SATVILab/UtilsFaustSV)
<!-- badges: end -->

The goal of UtilsFaustSV is to provide convenient functions to interrogate a FAUST clustering and extract output from it.

## Installation

You can install the package from [GitHub](https://github.com/) with:

``` r

if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("SATVILab/UtilsFaustSV")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(UtilsFaustSV)
```

First save the directory FAUST saved results to:
```{r }
# set the FAUST project path
proj_path <- usethis::proj_path("tests/testthat")
```

You can get the names of the markers FAUST used to cluster and the number of levels for each (after FAUST has clustered):
```{r }
get_faust_markers_and_levels(project_path = proj_path)
```

You can save FAUST-identified clusters as FCS files using only a subset of the markers (not run):
```{r , eval = FALSE}
save_faust_pop(
  project_path = proj_path, pop = list("CD3" = 2, "CD4" = 2),
  gs = gs
) # set gs to be the GatingSet FAUST clustered
```

You can get a table of counts of FAUST-identified clusters defined
using only a subset of the markers used (not run):
```{r , eval = FALSE}
get_pop_counts(
  project_path = proj_path,
  pop = c("CD4" = "-", "CD8" = "+")
)
```

You can save plots of counts of FAUST-identified clusters, where the clusters are defined using only a subset of the markers (not run):
```{r, eval = FALSE}
# plot all subsets that match these annotations individually
pop <- c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-")
plot_faust_count(
  project_path = proj_path,
  pop = pop
)

# plot total counts of subsets that match these annotations
# within each list element
pop <- list(
  c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-"),
  c("CD3" = "+", "CD4" = "-", "CD8-IgD" = "+")
)
plot_faust_count(
  project_path = proj_path,
  pop = pop
)
```
