---
title: "Getting Started with UtilsFaustSV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with UtilsFaustSV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

UtilsFaustSV provides utility functions to extract and work with outputs from
[FAUST](https://github.com/RGLab/FAUST) (Functional Annotation of Unsupervised
T-cell clustering) for use in downstream analyses.

The main use cases for this package are:

1. **Extracting FCS files** of cells belonging to FAUST-identified populations
   for further analysis in other tools
2. **Getting cell counts** for populations defined using a subset of the markers
   FAUST used for clustering
3. **Visualizing population frequencies** across samples

The key feature of this package is the ability to define populations using only
a subset of the markers that FAUST used. This allows you to:

- Extract broader cell lineages (e.g., all CD3+CD4+ cells regardless of other
  markers)
- Combine multiple FAUST clusters into a single population
- Get counts or FCS files for cell types defined at different levels of
  granularity

## Setup

```{r setup}
library(UtilsFaustSV)

# Set the path to your FAUST project directory
project_path <- "/path/to/faust/project"
```

## Understanding FAUST Output Structure

After running FAUST, the output is stored in a directory structure like:

```
project_path/
└── faustData/
    ├── faustCountMatrix.rds         # Count matrix
    ├── exhaustiveFaustCountMatrix.rds
    ├── metaData/
    │   ├── analysisMap.rds
    │   ├── scampClusterNames.rds
    │   └── activeChannels.rds
    ├── sampleData/
    │   └── <sample_name>/
    │       └── faustAnnotation.csv  # Per-cell annotations
    └── fcsData/                     # Output FCS files (created by this package)
```

## Getting Marker Information

First, it's useful to understand which markers FAUST used and how many expression
levels it identified for each:

```{r}
# Get markers and their levels
markers <- faust_marker_get_usage(project_path = project_path)
print(markers)
```

This returns a named numeric vector where:
- Names are the marker names
- Values are the number of expression levels (typically 2 for binary +/-)

## Defining Populations

Populations can be defined in several ways:

### Named Character Vector

The most common way to define a population is using a named character vector
where names are markers and values are expression levels:

```{r}
# CD3+CD4+ cells
pop_cd4 <- c("CD3" = "+", "CD4" = "+")

# CD3+CD8+ cells
pop_cd8 <- c("CD3" = "+", "CD4" = "-", "CD8-IgD" = "+")
```

### Expression Level Notation

For markers with more than 2 levels, you can use descriptive levels:

- `-`: Negative
- `+`: Positive (for 2-level markers)
- `Dim`: Dim expression (for 3-level markers)
- `Bright`: Bright expression (for 3-level markers)
- `MedLow`, `MedHigh`, `VeryBright`: For 4-level markers

### Numeric Notation

You can also use numeric levels (1 = lowest, 2 = next, etc.):

```{r}
# Equivalent to c("CD3" = "+", "CD4" = "+")
pop_cd4_numeric <- c("CD3" = 2, "CD4" = 2)
```

### Multiple Populations

Use a list to define multiple populations:

```{r}
# Both CD4+ and CD8+ T cells
pop_both <- list(
  c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-"),
  c("CD3" = "+", "CD4" = "-", "CD8-IgD" = "+")
)
```

## Writing FCS Files

The `faust_fcs_write()` function exports cells matching a population definition
to FCS files:

```{r}
# You need the original GatingSet that FAUST used
# gs <- flowWorkspace::load_gs("/path/to/gatingset")

# Export CD3+CD4+ cells
faust_fcs_write(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  fr_source = gs
)

# Files are saved to:
# project_path/faustData/fcsData/<pop_definition>/
```

### Custom Output Directory

```{r}
faust_fcs_write(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  fr_source = gs,
  dir_save = "/path/to/output"
)
```

### Exporting Specific Samples

```{r}
# By sample name
faust_fcs_write(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  fr_source = gs,
  sample = c("sample1.fcs", "sample2.fcs")
)

# By index
faust_fcs_write(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  fr_source = gs,
  sample = 1:5
)
```

### Back-transforming Data

If your data was transformed before FAUST analysis, you can back-transform
when exporting:
```{r}
# Inverse hyperbolic sine back-transformation
faust_fcs_write(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  fr_source = gs,
  trans_fn = sinh
)
```

## Getting Cell Counts

### Raw Count Matrix

Get the full FAUST count matrix:

```{r}
counts <- faust_count_get(project_path = project_path)
```

### Counts for Specific Populations

Get counts for populations matching a definition:

```{r}
# Get counts for all CD3+CD4+ subsets
cd4_counts <- faust_count_get_pop(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+")
)
```

The returned data frame includes:
- Sample identifiers
- Total counts and classified counts
- Counts for each matching population subset

### Summed Counts

When using a list of populations, counts are summed:

```{r}
# Get summed counts for T cell subsets
t_cell_counts <- faust_count_get_pop(
  project_path = project_path,
  pop = list(
    "CD4+ T cells" = c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-"),
    "CD8+ T cells" = c("CD3" = "+", "CD4" = "-", "CD8-IgD" = "+")
  )
)
```

### Exhaustive Count Matrix

By default, FAUST filters out rare populations. To include all populations:

```{r}
all_counts <- faust_count_get_pop(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  exhaustive = TRUE
)
```

## Plotting Population Counts

The `faust_count_plot()` function creates box plots of population frequencies:

```{r}
# Plot all CD4+ T cell subsets
faust_count_plot(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+", "CD8-IgD" = "-")
)
```

Plots are saved to:
`project_path/faustData/plotData/pop_stats/<population_name>.png`

### Customizing Plots

```{r}
faust_count_plot(
  project_path = project_path,
  pop = c("CD3" = "+", "CD4" = "+"),
  font_size = 12,
  point_size_max = 3,
  p_width = 50,
  p_height = 20
)
```

## Workflow Example

Here's a complete workflow for extracting and analyzing a T cell subset:

```{r}
library(UtilsFaustSV)

# Set paths
project_path <- "/path/to/faust/project"

# 1. Check available markers
markers <- faust_marker_get_usage(project_path = project_path)
print(markers)

# 2. Define populations of interest
cd4_memory <- c(
  "CD3" = "+",
  "CD4" = "+",
  "CD8-IgD" = "-",
  "CD45RA" = "-",
  "CCR7" = "+"
)

# 3. Get counts
counts <- faust_count_get_pop(
  project_path = project_path,
  pop = cd4_memory
)

# 4. Create visualization
faust_count_plot(
  project_path = project_path,
  pop = cd4_memory
)

# 5. Export FCS files for further analysis
faust_fcs_write(
  project_path = project_path,
  pop = cd4_memory,
  fr_source = gs  # Your GatingSet
)
```

## Session Info

```{r sessioninfo, eval = TRUE}
sessionInfo()
```
